A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-11-02 doc/large_strain.tex
most of Prof. Steinmann's comments

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the {\color{red}referential} configuration $\mcl B_0$ to the {\color{red}spatial} configuration $\mcl B_t$ at time $t$
is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
The current location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

%
% Deformation gradient
%
The tangent map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called deformation gradient and $\gz I$ is the second-order unit tensor.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
\begin{align}
  \chi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
  \chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}
Note that $\d \gz x = \chi(\d \gz X)$.

@@

The principle of stationary potential energy at equilibrium requires that the directional derivative with respect to the displacement
\begin{align}
\D_{\displaystyle \delta \gz u} \mcl E :=
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, \qquad \forall \delta \gz u \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
\left[\gz \tau \cdot \gz F^{-T}\right] : \Grad \, \delta \gz u =
\gz \tau : \left[\Grad \, \delta \gz u \cdot \gz F^{-1}\right] =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u) ,
\end{align}
where $\D_{\displaystyle \Delta \gz u}(\bullet)$ denotes the directional derivative in the direction $\Delta \gz u$.
For conservative traction boundary conditions, the directional derivative is given by
\begin{equation}
\begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
&=
\int_{\mcl B_0}
\D_{\displaystyle \Delta \gz u} \left(\gz F \cdot \gz S \cdot \gz F^T\right)  :
\overline{\grad^s} \, \delta \gz u
\, \d V
\\
& +
\int_{\mcl B_0}
\overline{\gz \tau} :
\left[
  \Grad \, \delta \gz u \cdot
  \D_{\displaystyle \Delta \gz u} \gz F^{-1}
\right] \d V.
\end{split}
\label{eq:tangent_pre}
\end{equation}
Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
are employed together with
$\D_{\displaystyle \Delta \gz u} \gz S = 2 \partial \gz S / \partial \gz C : \D_{\displaystyle \Delta \gz u} \gz E$.
}
\begin{equation}
  \begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  &=
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
  &+
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\end{split}
\label{eq:tangent}
\end{equation}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.

@@

For the current study, we use the Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated {\color{red}fourth-order material tangent tensor} are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}

@@

The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
\begin{align}
  A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} \equiv a(\gz N_i, \gz N_j) &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\displaystyle \tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}spatial} configuration, whereas the integration is done in the {\color{red}referential} configuration.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point on isoparametric element $\gz \xi_q$ and $J^K$ is the Jacobian of the mapping from the isoparametric element to the element $K$.
For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in that case that we consider a problem in $d$ space dimensions. For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

We briefly note that the SIMD vectorization is applied at the finite element level, that is the operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement the tangent {\color{red}operator} of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the {\color{red}spatial} configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the {\color{red}spatial} configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is performed over the {\color{red}spatial} configuration. Therefore, we have to divide by the Jacobian $J$ of the deformation map at a given quadrature point additionally to express the integral in the {\color{red}referential} configuration.

@@

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the {\color{red}referential} configuration.

@@

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the {\color{red}referential} configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material fourth-order tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

@@

Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent {\color{red}stiffness} tensor.
In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.

@@

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number of unknown DoFs $N$ is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorate for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.

>>>>

\subsection{Weak form}
%
% Kinematics
%
The deformation of a body $\mcl B$ from the {\color{red}referential} configuration $\mcl B_0$ to the {\color{red}spatial} configuration $\mcl B_t$ at time $t$
is defined via the deformation map $\gz \varphi_t: \mcl B_0 \rightarrow \mcl B_t$, which places material points of $\mcl B$ into the Euclidean space $\mathbb E^3$.
The {\color{red}spatial} location of a material particle $\gz X$ is given by $\gz x = \gz \varphi_t (\gz X)$.
The displacement field reads $\gz u = \gz x - \gz X$.

%
% Deformation gradient
%
The tangent map is linear such that
$\d \gz x = \gz F \cdot \d \gz X$,
where $\gz F := \Grad \, \gz \varphi \equiv \gz I + \Grad \, \gz u$ is called {\color{red}the} deformation gradient and $\gz I$ is the second-order unit tensor.
The mapping has to be one-to-one and must exclude self-penetration. Consequently, the Jacobian $J = \det \gz F > 0$ has to be positive.
% The deformation gradient can be split into the proper orthogonal rotation tensor $\gz R$ and the symmetric right $\gz U$ and left $\gz V$ stretch tensors such that $\gz F = \gz R \gz U = \gz V \gz R$.
We shall denote the push-forward transformation of rank-2 and rank-4 tensors $\mathbf{A}$ and $\boldsymbol{\mathcal{A}}$ by a linear transformation $\chi$ defined as
\begin{align}
  \chi\left( \mathbf{A} \right)_{ij}
  &= F_{iA} A_{AB} F_{jB} \\
  \chi\left( \boldsymbol{\mathcal{A}} \right)_{ijkl}
  &= F_{iA} F_{jB} \mathcal{A}_{ABCD} F_{kC} F_{lD}.
\end{align}
Note that $\d \gz x = \chi(\d \gz X)$.

@@

The principle of stationary potential energy at equilibrium requires that the directional derivative with respect to the displacement
\begin{align}
\D_{\displaystyle \delta \gz u} \mcl E :=
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, \qquad \forall \delta \gz u \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar{\color{red}-valued} non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
\left[\gz \tau \cdot \gz F^{-T}\right] : \Grad \, \delta \gz u =
\gz \tau : \left[\Grad \, \delta \gz u \cdot \gz F^{-1}\right] =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$
{\color{red}is involved due to the symmetry of $\gz \tau$}.
Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u) ,
\end{align}
where $\D_{\displaystyle \Delta \gz u}(\bullet)$ denotes the directional derivative in the direction $\Delta \gz u$.
For conservative traction boundary conditions, the directional derivative is given by
\begin{equation}
\begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
&=
\int_{\mcl B_0}
\D_{\displaystyle \Delta \gz u} \left(\gz F \cdot \gz S \cdot \gz F^T\right)  :
\overline{\grad^s} \, \delta \gz u
\, \d V
\\
& +
\int_{\mcl B_0}
\overline{\gz \tau} :
\left[
  \Grad \, \delta \gz u \cdot
  \D_{\displaystyle \Delta \gz u} \gz F^{-1}
\right] \d V.
\end{split}
\label{eq:tangent_pre}
\end{equation}
Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
are employed together with
$\D_{\displaystyle \Delta \gz u} \gz S = 2 \partial \gz S / \partial \gz C : \D_{\displaystyle \Delta \gz u} \gz E$.
}
\begin{equation}
  \begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  &=
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
  &+
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\end{split}
\label{eq:tangent}
\end{equation}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order {\color{red}spatial} tangent {\color{red}stiffness} tensor is the push forward of the material {\color{red}part of the referential} tangent {\color{red}stiffness tensor} $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$.
The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and {\color{red}is} therefore called material part of the tangent.
The second term in \eqref{eq:tangent} is called {\color{red}the} geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.

@@

For the current study, we use the Neo-Hookean model
\begin{gather}
\psi \left( \mathbf{C} \right)
  = \frac{\mu}{2} \left[ \trace{\mathbf{C}} - \trace{\mathbf{I}} - 2 \ln\left( J \right) \right]
  + \lambda \ln^{2}\left( J \right)
\end{gather}
where $\mu$ and $\lambda$ denote the shear modulus and Lam\'{e} parameter respectively. It can be shown that the first and second derivatives of the strain energy function are given by
\begin{align}
\frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}}
  &= \frac{\mu}{2} \mathbf{I} - \frac{1}{2} \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{C}^{-1} \\
\frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}}
  &= \frac{1}{2}\left[ \mu - 2\lambda\ln\left( J \right) \right] \left[ - \frac{d \mathbf{C}^{-1}}{d \mathbf{C}} \right]
  + \frac{\lambda}{2} \mathbf{C}^{-1} \otimes \mathbf{C}^{-1}
\end{align}
The Kirchhoff stress and its associated {\color{red}fourth-order material part of spatial tangent tensor} are
\begin{gather}
\boldsymbol{\tau}
  \equiv J \boldsymbol{\sigma}
  = \chi\left( 2 \frac{d \psi \left( \mathbf{C} \right)}{d \mathbf{C}} \right)
  = \mu \mathbf{b} - \left[ \mu - 2\lambda\ln\left( J \right) \right] \mathbf{I}
\end{gather}
and
\begin{gather}
J \boldsymbol{\mathcal{C}}
  = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \boldsymbol{\mathcal{S}}
  + 2 \lambda \mathbf{I} \otimes \mathbf{I}
\end{gather}
where $\boldsymbol{\mathcal{S}}$ is the fourth-order symmetric identity tensor.
The action that $J \boldsymbol{\mathcal{C}}$ performs when contracted with an arbitrary rank-2 symmetric tensor is therefore
\begin{gather}
J \boldsymbol{\mathcal{C}} : \left( \bullet \right)
  = 2 \left[ \mu - 2\lambda\ln\left( J \right) \right] \left( \bullet \right)
  + 2 \lambda \trace\left( \bullet \right) \mathbf{I}.
\label{eq:simplified_action}
\end{gather}

@@

The Newton-Raphson solution approach is adopted for the nonlinear problem considered here.
Given the current trial solution field $\overline{\gz u^h}$, the correction $\Delta \gz u^h$ field is obtained as the solution to the following system of equations (see \eqref{eq:weak_form} and \eqref{eq:tangent}):
\begin{align}
  \sum_{j \in \mcl I} A_{ij} \Delta u_j &= - F_i  \label{eq:linear_system} \\
  A_{ij} \equiv a(\gz N_i, \gz N_j) &=
  \int_{\mcl B_0}
  \underbrace{
  \overline{\grad^s} \gz N_i : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \gz N_j
  +
  \overline{\grad}\gz N_i :
  \left[
  \overline{\grad} \gz N_j \cdot
  \overline{\gz \tau}
  \right]
  }_{\displaystyle \tilde{a}(\gz N_i, \gz N_j)}
  \d V
  \label{eq:algebraic_tangent}
  \\
  F_i &=
  \int_{\mcl B_0} \overline{\gz \tau} : \grad^{s} \, \gz N_i \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \gz N_i \, \d S.
\end{align}
Here $\gz A$ is the discrete tangent {\color{red}operator}, $\gz F$ is the discrete gradient {\color{red}of the potential energy}
and $\tilde{a}(\gz N_i, \gz N_j)$ is a representation for the integrand in the bilinear form $a(\gz N_i, \gz N_j)$ for the trial solution field $\overline{\gz u^h}$.
Note that $\overline{\grad^s} \gz N_i$ and $\overline{\grad} \gz N_i$ are gradients of shape functions in the {\color{red}spatial} configuration, whereas the integration is done in the {\color{red}referential} configuration.

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the {\color{red}residual} force vector $\gz F$ corresponding to the discretization are assembled and
a direct or iterative solver is used to solve the linear system.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches. The idea is to perform the operations within the solver on-the-fly
rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ {\color{red}and
$w_q$ is the quadrature weight}.
For the sake of demonstration, let us assume that $\gz A$ represents a mass matrix with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ {\color{red}in $d$ space dimensions}.
For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

We briefly note that the SIMD vectorization is applied at the finite element level, {\color{red}i.e.} the {\color{red}matrix-free} operator is applied simultaneously on several elements (called ``blocks'') whereas the shared memory parallelization using Intel Threading Building Blocks is adopted at the element blocks.

Given that the matrix-free operators are applied on-the-fly, it is crucial to find a good balance between pre-calculation of certain values (caching) and performance of the algorithm.
Below, we propose three algorithms to implement the tangent {\color{red}operator} of finite-strain elasticity (see \eqref{eq:algebraic_tangent}) using the matrix-free approach.
%
In order to obtain gradients with respect to the {\color{red}spatial} configuration we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the {\color{red}spatial} configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is {\color{red}eventually} performed over the {\color{red}spatial} configuration.
Therefore, we have to {\color{red}additionally} divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the {\color{red}referential} configuration.

@@

Algorithm \ref{alg:mf_scalar} caches a single scalar which involves the logarithm of the Jacobian -- a computationally demanding operation compared to additions and multiplications. As evident from Section \ref{sec:theory}, we need to re-evaluate the Kirchhoff stress at each quadrature point in order to apply the {\color{red}tangent} operator. This in turn requires evaluating the deformation gradient $\gz F$ and therefore gradients with respect to the {\color{red}referential} configuration.

@@

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the {\color{red}referential} configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor} $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

@@

Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the {\color{red}material part of the fourth-order spatial tangent stiffness tensor}.
In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.

@@

Efficient and scalable matrix-vector products are not enough to obtain a method suitable for large computations. In particular, linear scaling with respect to the number $N$ of unknown DoFs is desirable.
The efficiency of standard iterative solvers and preconditioners deteriorates for large systems of linear equations due to increasing iteration counts under mesh refinement.
In contrast to other preconditioners, geometric multigrid (GMG) preconditioners \cite{Bramble1990, Briggs2000, Janssen2011,May2015} can be proven to result in iteration counts that are independent of the number of mesh refinements by smoothing the residual on a hierarchy of meshes. For further information about geometric multigrid methods we refer the reader to \cite{Briggs2000,Hackbusch1985,Wesseling1992}.
